---
title: "TCGA Head and Neck Cancer Example analysis"
subtitle: "DNA methytion predicted proteins, tissue type, and cancer progression"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: Author name
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
execute:
  cache: true
---

## Load cleaned TCGA clinical and predicted protein level data

```{r globals, results = 'hide'}
#| label: setup
#| include: false

library(ggplot2)
library(ggrepel)
library(meffonym)
library(here)

readRenviron(here("config.env"))
datadir <- Sys.getenv("datadir")
resultsdir <- Sys.getenv("resultsdir")
```

```{r load.data}
combined.filename <- file.path(resultsdir, "combined-clin-pred-proteins.txt")
data <- read.table(combined.filename,
  header = T, sep = "\t", stringsAsFactors = F
)
```

```{r names, results = 'hide'}
protein.names <-
  subset(
    meffonym::meffonym.models(full = T),
    grepl("^episcores", filename)
  )$name
protein.names <- make.names(protein.names)

table(protein.names %in% colnames(data))
```

## Run glms for association between proteins levels and tissue type (tumor vs. normal)

```{r tissue}
## define glm formulae with pred.proteins as predictors of 'tumor.or.normal'
## tissue i.e. tumor.or.normal ~ pred.protein
formulae <- sapply(protein.names, function(i) {
  reformulate(i, response = "tumor")
}, simplify = F)

# run glms
fit <- sapply(formulae, function(i) {
  glm(i, data = data, family = binomial())
}, simplify = F)

fit.summary <- sapply(fit, function(i) {
  out <- summary(i)$coefficients
  out[, "Estimate"] <- out[, "Estimate"]
  out
}, simplify = F)

fit.coefs <- sapply(fit.summary, function(i) {
  i[2, c("Estimate", "Pr(>|z|)")]
}, simplify = F)
fit.coefs <- {
  x <- do.call(rbind, fit.coefs)
  data.frame(
    pred.protein = rownames(x),
    coef = x[, "Estimate"],
    p.value = x[, "Pr(>|z|)"]
  )
}

bonferroni <- -log10(0.05 / length(fit))
```

### Visualize results

```{r tissue.figs}
fit.coefs |>
  ggplot(aes(x = pred.protein, y = -log10(p.value))) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  geom_text_repel(
    data = fit.coefs[which((-log10(fit.coefs$p.value) > bonferroni)), ],
    aes(x = pred.protein, y = -log10(p.value), label = pred.protein)
  ) +
  geom_hline(
    yintercept = bonferroni,
    linetype = "dashed"
  )

fit.coefs |>
  ggplot(aes(x = coef, y = -log10(p.value))) +
  geom_point() +
  geom_text_repel(
    data = fit.coefs[which((-log10(fit.coefs$p.value) > bonferroni)), ],
    aes(x = coef, y = -log10(p.value), label = pred.protein)
  ) +
  geom_hline(
    yintercept = bonferroni,
    linetype = "dashed"
  )

```

## Run glms for association between proteins levels and progression free interval (PFI)

This analysis should be restricted to measurements taken from tumor samples

```{r pfi}
tumor.data <- subset(data, tumor == 1)

## define glm formulae with pred.proteins as predictors of pfi
## tissue i.e. pfi ~ pred.protein
formulae <- sapply(protein.names, function(i) {
  reformulate(i, response = "pfi")
}, simplify = F)

# run glms
fit <- sapply(formulae, function(i) {
  glm(i, data = tumor.data, family = binomial())
}, simplify = F)

fit.summary <- sapply(fit, function(i) {
  out <- summary(i)$coefficients
  out[, "Estimate"] <- out[, "Estimate"]
  out
}, simplify = F)

fit.coefs <- sapply(fit.summary, function(i) {
  i[2, c("Estimate", "Pr(>|z|)")]
}, simplify = F)
fit.coefs <- {
  x <- do.call(rbind, fit.coefs)
  data.frame(
    pred.protein = rownames(x),
    coef = x[, "Estimate"],
    p.value = x[, "Pr(>|z|)"]
  )
}
```

### Visualize results

```{r pfi.figs}
fit.coefs |>
  ggplot(aes(x = pred.protein, y = -log10(p.value))) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  geom_text_repel(
    data = fit.coefs[which((-log10(fit.coefs$p.value) > bonferroni)), ],
    aes(x = pred.protein, y = -log10(p.value), label = pred.protein)
  ) +
  geom_hline(
    yintercept = bonferroni,
    linetype = "dashed"
  )

fit.coefs |>
  ggplot(aes(x = coef, y = -log10(p.value))) +
  geom_point() +
  geom_text_repel(
    data = fit.coefs[which((-log10(fit.coefs$p.value) > bonferroni)), ],
    aes(x = coef, y = -log10(p.value), label = pred.protein)
  ) +
  geom_hline(
    yintercept = bonferroni,
    linetype = "dashed"
  )
```


## Table of results

```{r}
library(gt)
library(gtsummary)
# Summary table
t1 <- tbl_summary(select(data, 
                         c(female, 
                           histology, 
                           age.at.diagnosis,
                           ethnicity,
                           race,
                           stage,
                           tumor.or.normal)), 
                  missing = "no",
                  label = list(
                    female = "Female",
                    histology = "Histology",
                    age.at.diagnosis = "Age at diagnosis",
                    ethnicity = "Ethnicity",
                    race = "Race",
                    stage = "Stage"), 
                  by = tumor.or.normal) |> 
  modify_caption("**Table 1: Cohort Characteristics**") |>
  bold_labels() |>
  as_gt() |>
  gt::cols_width(
    label ~ gt::px(250),    # Set first column to 250px
    everything() ~ gt::px(150))  # Set other columns to 150px

t1
```